Apparent preferences for cognitive effort fade when multiple forms of effort

### 2021-05-30 Manuscript.Rmd
done with discussion edits

These choice dynamics ruled out a number of common explanatory candidates. For instance, if the subjective costs of effort or delay were intrinsic, we would not have observed the gradual convergence in acceptances when they were all intermixed in the same environment. This also speaks against a direct role of boredom in our results. The aversiveness of boredom can lead people to attempt any available activity in order to avoid it [@westgate2020]. This would have likely led to a persistent preference for either effortful condition. Lastly, this gradual convergence in choices rules out the possibility that our sample was mostly composed of demand seekers. Some individuals express a strong need for cognition, which can bias studies of mental labor [@Cacioppo1996; @Sayali2019]. Even though the prevalence of this trait in our college-level population could have driven the preferences for cognitive effort observed in both experiments, we would have expected such individuals to maintain this preference in the face of alternatives. Another unlikely candidate is fatigue. Even though participants decreased their acceptance rates for effort in both experiments, we would have expected them to increase engagement with wait trials over time in order to preserve earning rates. Further, we do not expect any of our observations to be due to the risk of failing to successfully complete effortful trials, as participants remained accurate throughout both experiments. Finally, we cannot conclude that cognitive effort gained value from its rewarding outcomes, as has been proposed in the past [@Eisenberger1996]. This phenomenon could not have benefitted a single type of demand, as cognitive effort, physical effort, and unfilled delay were rewarded at the same rate.

  Traditional foraging models, in which a reward is weighed against the earnings foregone by pursuing it [@Charnov1976; @Constantino2015], were also not enough to explain our results. Instead, our results were consistent with a computational model in which each type of demand distinctly biased the estimation of the opportunity cost, and such biases could be gradually recalibrated over time as decision makers gained experience with additional demands. In this formulation, passive delay and physical effort added to the opportunity cost, whereas cognitive effort reduced it. This raises the question of how these biases arise. We hypothesize that demands altered the perceived duration of time. Immersive activities can subjectively speed up the passage of time [@Csikszentmihalyi2014], and this effect appears to scale with the amount of cognitive processing required by a task [@Wearden2016]. Moreover, adjusting for time sensitivity has been shown to improve fits for delay discounting models [@Mckerchar2009]. Therefore, even though participants knew the nominal handling times, cognitive effort might have become valuable by subjectively compressing the perceived durations. In our experiments, shorter subjective handling times would result in higher perceived earning rates, but it is also possible that individuals benefitted from a sense of fast completion. We speculate that this construal of elapsed time could then be adjusted with the experience of alternative demands, as participants learned how long the 10 seconds handling time felt. This could help explain why preferences for cognitive effort were smaller for within-subject participants who started with the physical block, as they had already formed an estimate of the handling time by the time they experienced a subjectively faster cognitive trial. The shorter time sensitivity produced by cognitive effort could be due to participants counting the number of sub-tasks during pursual (which was always less than the number of seconds in the handling time), or because taxing cognitive resources distracted individuals from estimating the passage of time [@Zakay1989]. These hypotheses will have to be experimentally probed in the future.

  Relying on foraging paradigms helped us to control for common shortfalls in the existing literature. For example, even though engaging with a demanding offer encompasses evaluating, pursuing, and receiving its outcome, investigations of the pursuit and receipt stages have been limited. Many studies have sought to isolate the evaluation stage by presenting individuals with offers that do not come into effect until the end of the experimental session, deliberately setting aside the question of how the ongoing experience of cost and outcome affects subsequent choices [@Kable2007; @Chong2016; @Massar2015; @Apps2015a]. However, discrepancies between expected and experienced value are well known [@Kahneman2006], and choosing immediate outcomes could be due to factors other than time preference, such as the risk of being unable to redeem the future rewards, or hoping to find better offers in the near future [@Fawcett2012; @Frederick2002; @Sozou1998]. We circumvented these pitfalls by providing participants with an opportunity to actively experience the demands they accepted, and giving them full knowledge of the available options. Another challenge in comparing effort and delay is that increments in the required effort are often paired with a longer execution time. This issue is often avoided by comparing discounting functions, which yield points at which individuals faced with increasing costs become indifferent to large rewards [@Massar2015; @Prevost2010]. We eliminated this possibility by fixing the handling time for all demands, allowing us to probe people's preferences directly. Moreover, Experiment 1 provided a baseline of foraging behavior that contextualized the findings of the second experiment (a direct test of our hypothesis). Finally, the existence of optimal strategies helped us identify when demands could be attractive, a central motivation for this study. The apparent attractiveness of cognitive effort was reflected both by a higher proportion of acceptances at the expense of earnings, and by the fitted opportunity costs biases discussed above. Of note is that the optimality of the strategies induced by each demand depended on the richness of the block. This can help explain why the wait group earned the most despite having similar acceptance rates to the physical group, as the former was closer to optimality in the richest environment.

  A potential limitation of this work lies in whether our task manipulations were actually costly. For instance, it is possible that participants gravitated towards cognitive demands because the task configuration was not effortful. This could be due to differences in how we calibrated cognitive and physical tasks. For physical demands, we adjusted the gripping threshold to people's maximum strength, whereas we trained participants in the cognitive tasks until they reached criterion. Training might have improved response automaticity and reduced error-related risk. In addition, attributes such as the efficacy to complete a task, or the diversity of its components, have been reported to promote engagement [@Bandura2010; @Eisenberger1996]. However, hand grips are regularly used to probe physical effort [@Chong2017; @Hogan2018; @Prevost2010], and task switching has been shown to feel cognitively effortful by recruiting cognitive control [@Kool2010; @Shenhav2017]. Furthermore, we observed a steeper decrease in acceptances over time for both effort domains, a hallmark of effortful exertion [@Treadway2009].

  The present findings have important implications for daily and clinical scenarios. Excessive aversion to delay and effort can manifest as impulsivity and apathy, respectively [@Ainslie1975; @Bonnelle2016; @Chong2016; @Cummings1993], fluctuations in which are hallmark symptoms of various mental illnesses and substance use disorders [@Paulus2007]. Lack of motivation to exert effort or wait for prospects is a distinguishing factor among mental afflictions like depression and schizophrenia [@Barch2019]. Within the former, medicated patients in remission are more sensitive to effort than controls, and this sensitivity is predictive of relapse [@Berwian2020]. Encouraging patients to engage in activities has been found to be a helpful treatment for depression [@Dimidjian2006]. Tailored reward schedules can successfully achieve this [@Eisenberger1996]. However, since depressed patients also show insensitivity to rewards [@Huys2013], identifying what makes effort appealing in its own right can help breach treatment-resistant cases. Our work suggests that subdividing a task can promote its pursuit, although further testing is required. More importantly, it highlights that decision scenarios should be configured to promote pursuit of desired activities, perhaps by removing alternative uses of time. These same principles can potentially be applied to increasing productivity at work or adherence to physical exercise routines.

@@

  In this study, we have examined the factors that motivate engagement with different behavioral demands, emphasizing how these can be molded by the decision maker's environment. Unlike existing theories that have focused on the role of incentives and retrospective justifications [@Eisenberger1996; @Inzlicht2018], our work highlights outcome-independent and task-specific features that can motivate the pursuit of effortful avenues, which can be potentially manipulated in order to promote engagement in clinical and daily activities.

>>>>

  The observed choice dynamics ruled out a number of candidate explanations. For instance, if the subjective costs and benefits of effort or delay were intrinsic, we would not have expected to observe the gradual convergence in acceptances when all three forms of demand were intermixed in the same environment. This line of reasoning also speaks against a direct role of boredom in our results. The aversiveness of boredom can lead people to attempt any available activity in order to avoid it [@westgate2020]. In our paradigm, such a tendency would have been expected to lead to a persistent preference for either effortful condition. Lastly, the gradual convergence of acceptance rates across cost conditions  rules out the possibility that our sample was mostly composed of demand seekers. Some individuals express a strong need for cognition, which can bias studies of mental labor [@Cacioppo1996; @Sayali2019]. Even though the prevalence of this trait in our college-level population could have driven the preferences for cognitive effort observed in both experiments, we would have expected such individuals to maintain this preference in the face of alternatives. Another unlikely candidate is fatigue. Even though participants decreased their acceptance rates for effort in both experiments, we would have expected participants in Experiment 2 to increase engagement with wait trials, substituting wait trials for effort trials in order to preserve their earnings. Our observations are unlikely to be due to the risk of failing to successfully complete effortful trials, as participants remained accurate throughout both experiments. Finally, we cannot conclude that cognitive effort gained value from its rewarding outcomes, as has been proposed in the past [@Eisenberger1996]. This phenomenon would not have been expected to benegit a single type of demand differentially because cognitive effort, physical effort, and unfilled delay were rewarded at the same rate.

  Traditional foraging models, in which a reward is weighed against the earnings foregone by pursuing it [@Charnov1976; @Constantino2015], were also not sufficient to explain our results. Instead, our results were consistent with a computational model in which each type of demand distinctly biased the estimation of the opportunity cost, and such biases could be gradually recalibrated over time as decision makers gained experience with additional demands. In this formulation, passive delay and physical effort added to the opportunity cost, whereas cognitive effort reduced it. This raises the question of how these biases arise. We hypothesize that demands altered the perceived duration of the delay interval. Immersive activities can subjectively speed up the passage of time [@Csikszentmihalyi2014], and this effect appears to scale with the amount of cognitive processing required by a task [@Wearden2016]. Moreover, adjusting for time sensitivity has been shown to improve fits for delay discounting models [@Mckerchar2009]. Therefore, even though participants knew the nominal handling times, cognitive effort might have promoted higher acceptance rates by subjectively compressing the perceived durations. We speculate that this construal of elapsed time could then be adjusted with the experience of alternative demands, as participants learned how long the 10-second handling time felt in the context of various activities. This could help explain why preferences for cognitive effort were smaller for those participants in Experiment 2 who started with the physical effort block, as they had already begun to form an estimate of the handling time before they experienced their first cognitive effort trial. The cognitive effort task might have shortened subjective durations for several reasons, either because it divided the handling time into discrete, 2 s trials, or because taxing cognitive resources distracted individuals from estimating the passage of time [@Zakay1989]. These hypotheses will have to be experimentally probed in the future.

  Relying on foraging paradigms enabled us to address several gaps in the existing literature of effort-based decision making. For example, even though engaging with a demanding offer encompasses the evaluation, pursuit, and receipt of the outcome, investigations of the pursuit and receipt stages have been limited. Many studies have sought to isolate the evaluation stage by presenting individuals with offers that are not realized until the end of the experimental session, deliberately setting aside the question of how the ongoing experience of cost and outcome affects subsequent choices [@Kable2007; @Chong2016; @Massar2015; @Apps2015a]. However, discrepancies between expected and experienced value are well known [@Kahneman2006], and choosing immediate or low-effort outcomes could be due to factors other than time preference, such as a perceived risk of being unable to redeem the future rewards, or hoping to find better offers in the near future [@Fawcett2012; @Frederick2002; @Sozou1998]. We circumvented these pitfalls by providing participants with an opportunity to actively experience the demands they accepted, and giving them full knowledge of the available options. Another challenge in comparing effort and delay is that increments in the required effort are often paired with a longer execution time. We avoided this issue by imposing consistent handling times for all demands, allowing us manipulate temporal durations and demands for effort independently. Finally, the existence of well-defined optimal strategies provided a useful reference point for evaluating choice behavior. The apparent attractiveness of cognitive effort was reflected both by a higher proportion of acceptances at the expense of earnings, and by parameter estimates capturing bias in the estimation of opportunity costs.

  A potential limitation of this work is that we examined only a single combination of cognitively effortful tasks. Future work should investigate whether other ways of operationalizing cognitive and physical effort would be more aversive or more engaging and would resultingly have different effects on decision making. A potentially relevant factor is the way in which we calibrated cognitive and physical tasks. For physical demands, we adjusted the gripping threshold to people's maximum strength, whereas we trained participants in the cognitive tasks until they reached criterion. Training might have improved response automaticity and reduced error-related risk. In addition, attributes such as the efficacy to complete a task, or the diversity of its components, have been reported to promote engagement [@Bandura2010; @Eisenberger1996]. However, hand grip tasks have frequently been used to probe physical effort in previous work [@Chong2017; @Hogan2018; @Prevost2010], and task switching has been shown to impose subjective costs [@Kool2010; @Shenhav2017]. Furthermore, the observed decrease in acceptance rates over time for both effort domains is consistent with patterns seen for other forms of effortful exertion [@Treadway2009].

  The present findings have important implications for everyday and clinical scenarios. Excessive aversion to delay and effort can manifest as impulsivity and apathy, respectively [@Ainslie1975; @Bonnelle2016; @Chong2016; @Cummings1993], fluctuations in which are hallmark symptoms of various mental illnesses and substance use disorders [@Paulus2007]. Lack of motivation to exert effort or wait for delayed prospects is a distinguishing factor of conditions such as depression and schizophrenia [@Barch2019]. In depression, medicated patients in remission are more sensitive to effort than control participants, and this sensitivity is predictive of relapse [@Berwian2020]. Encouraging patients to engage in activities has been found to be a helpful treatment for depression [@Dimidjian2006]. Tailored reward schedules can successfully achieve this [@Eisenberger1996]. However, since depressed patients also show insensitivity to rewards [@Huys2013], identifying what makes effort appealing in its own right can help address treatment-resistant cases. Our work suggests that voluntary effortful engagement can be encouraged by removing cognitively less-effortful paths to reward from the immediate decision context, and suggests it would be useful for future work to test directly for effects of effortful tasks on concurrent time estimation. These same principles can potentially be applied to increasing productivity at work or adherence to physical exercise routines.

@@

  In this study, we have examined the factors that motivate engagement with different behavioral demands, emphasizing how engagement can be influenced by features of the decision environment. Complementing work that has focused on the role of incentives and retrospective justifications [@Eisenberger1996; @Inzlicht2018], our work highlights outcome-independent and task-specific features that can motivate the pursuit of effortful courses of action.
